### 2014-03-31 ApTranscriptome_TR.Rmd
Changed from robust regression, to simply using log-transformed response to control for outliers
* Reworking analysis script to save results of `lm` for expression of each transcript in a list.
* Fit model using log to lessen effects of outliers with high expression on results. This had little effect, actually slightly increasing the number of significant transcripts.

**March 20, 2014**

@@

# Load libraries
library(R.utils)
library(ggplot2)
library(knitr)
library(knitcitations)
library(pander)
library(stringr)
library(data.table)
library(RCurl)
library(plyr)

@@

setkey(TPM.dt, val)
TPM.dt.sub <- TPM.dt[val != 7] 
unique(TPM.dt.sub$val)

# define model for RxN function
model <- "TPM ~ colony + val + I(val^2) + colony:val + colony:I(val^2)"

# identify responsive transcripts
RxNout <- RxNseq(f = TPM.dt.sub, model = model, threshold = 0.05)

save.image("RxN_combined_results.RData")
```

Of the `r nrow(RxNout)` transcripts, `r length(which(RxNout$pval < 0.05))` have models with P < 0.05.

Many of these are likely false positives, so I adjusted P-values using false discovery rate (FDR) to identify only those transcripts with less than 5% FDR as significant.

```{r fdr}
RxNout$padj <- p.adjust(RxNout$pval, method = "fdr")
# Plot FDR values against initial pvalues
par(mfrow = c(2,1))
hist(RxNout$pval)
hist(RxNout$padj)

# subset to significant transcripts
signif.transcripts <- RxNout[which(RxNout$padj < 0.05), ]
```

@@

```{r responsive_transcripts}
# transcripts that differ in expression by colony only
colony.transcripts <- signif.transcripts[!is.na(signif.transcripts$'coef.colony') & is.na(signif.transcripts$'coef.val') & is.na(signif.transcripts$'coef.I(val^2)') & is.na(signif.transcripts$'coef.colony:val') & is.na(signif.transcripts$'coef.colony:I(val^2)')]
# check that subset correct - no transcripts with significant temperature terms
length(which(!is.na(colony.transcripts$'coef.val') | !is.na(colony.transcripts$'coef.I(val^2)') | !is.na(colony.transcripts$'coef.colony:val') | !is.na(colony.transcripts$'coef.I(val^2)')))
str(colony.transcripts)
# order by adjusted P-value
colony.transcripts <- colony.transcripts[order(colony.transcripts$padj), ]

# remaining transcripts are thermally responsive
responsive.transcripts <- signif.transcripts[!is.na(signif.transcripts$'coef.val') | !is.na(signif.transcripts$'coef.I(val^2)') | !is.na(signif.transcripts$'coef.colony:val') | !is.na(signif.transcripts$'coef.colony:I(val^2)')]
dim(responsive.transcripts)

# check that these two groups sum to total significant transcripts
stopifnot(all.equal(nrow(signif.transcripts), nrow(colony.transcripts) + nrow(responsive.transcripts)))

# split responsive_transcripts into those that have same temperature response in both colonies and 
# main effect of temperature only
temperature.transcripts <- responsive.transcripts[is.na(responsive.transcripts$'coef.colony:val') & is.na(responsive.transcripts$'coef.colony:I(val^2)')]
# order by adjusted P-value
temperature.transcripts <- temperature.transcripts[order(temperature.transcripts$padj), ]
str(temperature.transcripts)

# transcripts that have colony x temperature interaction
interaction.transcripts <- signif.transcripts[!is.na(signif.transcripts$'coef.colony:val') | !is.na(signif.transcripts$'coef.colony:I(val^2)')]
# order by adjusted P-value
interaction.transcripts <- interaction.transcripts[order(interaction.transcripts$padj), ]
str(interaction.transcripts)

# check that these two groups sum to total responsive transcripts
stopifnot(all.equal(nrow(responsive.transcripts), nrow(temperature.transcripts) + nrow(interaction.transcripts)))
```

```{r table_results, echo=FALSE, results='asis'}
clist <- c('Total', 'Colony', 'Temperature', 'Temperature:Colony')
siglist <- c(nrow(signif.transcripts),
             nrow(colony.transcripts),
             nrow(temperature.transcripts),
             nrow(interaction.transcripts)
             )

>>>>

**March 31, 2014**

@@

# Load libraries
library(R.utils)
library(ggplot2)
library(knitr)
library(knitcitations)
library(pander)
library(stringr)
library(data.table)
library(RCurl)
library(plyr)
library(MASS)

@@

setkey(TPM.dt, val)
TPM.dt.sub <- TPM.dt[val != 7] 
unique(TPM.dt.sub$val)
```

(1) Identify transcripts with overall significant model fit. Adjust *P* values for multiple testing using FDR and retain transcripts with FDR < 0.05. Use log-transformed response to account for outliers.

```{r modp}
# define model for RxN function
model <-  "log(TPM+1) ~ colony + val + I(val^2) + colony:val + colony:I(val^2)"

# define function for `ddply` to calculate overall model P value
modpFunc <- function(foo) {
  # fit lm
  lmout <- eval(parse(text = paste("lm(", model, ", data = foo)", sep = "")))

  # calculate overall model significance
  f <- summary(lmout)$fstatistic
  pval <- unname(pf(f[1],f[2],f[3],lower.tail=F))
  attributes(pval) <- NULL

  # return pvalue
  c(pval = pval) 
}

RxNpval <- ddply(TPM.dt.sub, .(Transcript), .progress="text", .inform="TRUE", modpFunc)
```

Of the `r nrow(RxNpval)` transcripts, `r length(which(RxNpval$pval < 0.05))` have models with P < 0.05.

Many of these are likely false positives, so I adjust P-values using false discovery rate (FDR). Only those transcripts with less than 5% FDR are retained as significant.

```{r fdr}
RxNpval$padj <- p.adjust(RxNpval$pval, method = "fdr")
# Plot FDR values against initial pvalues
par(mfrow = c(2,1))
hist(RxNpval$pval)
hist(RxNpval$padj)

# subset to significant transcripts
signif.transcripts <- RxNpval[which(RxNpval$padj < 0.05), ]

# extract significant transcripts
sig.TPM.dt.sub <- TPM.dt.sub[signif.transcripts$Transcript]
```

At the 5% FDR significance threshold, there are `r nrow(signif.transcripts)` transcripts with an overall significant model.

(2) Fit linear model to overall significant transcripts; perform stepAIC to retain only significant terms, and save `lm` output to list

```{r lmFunc}
# first, define function to use
lmFunc <- function(bar) {
  # fit lm
  lmout <- eval(parse(text = paste("lm(", model, ", data = bar)", sep = "")))

  # stepAIC to drop non-significant terms
  f.lmout <- stepAIC(lmout)

  # return final model
  return(f.lmout) 
}

# apply function to all responsive transcripts
RxNlmAIC <- dlply(sig.TPM.dt.sub, .(Transcript), .progress="text", lmFunc)
```

@@

```{r responsive_transcripts}
# transcripts that differ in expression by colony only
keepthis <- function(lmitem, term = NA) {
  coefs <- coefficients(lmitem)
  coef.names <- names(coefs)
  keep <- ifelse(term %in% coef.names, TRUE, FALSE)
}

# identify sets of transcripts that include each term in final model
colony.transcripts <- RxNlmAIC[which(Map(keepthis, RxNlmAIC, term = "colonyAr") == TRUE)]
temp.lin.transcripts <- RxNlmAIC[which(Map(keepthis, RxNlmAIC, term = "val") == TRUE)]
temp.quad.transcripts <- RxNlmAIC[which(Map(keepthis, RxNlmAIC, term = "I(val^2)") == TRUE)]
col.by.temp.transcripts <- RxNlmAIC[which(Map(keepthis, RxNlmAIC, term = "colonyAr:val") == TRUE)]
col.by.temp.quad.transcripts <- RxNlmAIC[which(Map(keepthis, RxNlmAIC, term = "colonyAr:I(val^2)") == TRUE)]

# union of thermally-responsive transcripts
responsive.transcripts <- union(temp.lin.transcripts, temp.quad.transcripts)
responsive.transcripts <- union(responsive.transcripts, col.by.temp.transcripts)
responsive.transcripts <- union(responsive.transcripts, col.by.temp.quad.transcripts)
length(responsive.transcripts)

# transcripts that have a colony effect only 
colony.transcripts <- setdiff(RxNlmAIC, responsive.transcripts)

# transcripts that have colony x temperature interaction
interaction.transcripts <- union(col.by.temp.transcripts, col.by.temp.quad.transcripts)

# split responsive_transcripts into those that have same temperature response in both colonies and 
# main effect of temperature only
temperature.transcripts <- setdiff(responsive.transcripts, interaction.transcripts)
```

```{r table_results, echo=FALSE, results='asis'}
clist <- c('Total', 'Colony', 'Temperature', 'Temperature:Colony')
siglist <- c(length(RxNlmAIC),
             length(colony.transcripts),
             length(temperature.transcripts),
             length(interaction.transcripts)
             )
